---
title: "Richness and endemic species by EEZ"
editor_options: 
  chunk_output_type: console
---

The goal of this notebook is to find species which have been detected in only one EEZ according to OBIS.

The code for this notebook is available at <https://github.com/iobis/notebook-endemics>.

## Species by EEZ

Here we generate species lists by EEZ. We use a dataset of precalculated H3 grid cells by EEZ, which is based on the [Marine Regions](https://marineregions.org/) EEZ layer.

```{r, message=FALSE, warning=FALSE, cache=TRUE}
library(arrow)
library(dplyr)
library(h3)
library(h3jsr)

h3_res <- 7

# aggregate OBIS data by H3 cell and species

row_to_geo <- function(row, res) {
  geo_to_h3(c(row$decimalLatitude, row$decimalLongitude), res)
}

gridded_species <- open_dataset("~/Desktop/temp/obis_20230726.parquet") %>%
  group_by(decimalLongitude, decimalLatitude, speciesid, species) %>%
  summarize() %>%
  collect() %>%
  as.data.frame() %>%
  mutate(h3 = row_to_geo(., h3_res)) %>%
  group_by(speciesid, species, h3) %>%
  summarize() 

# read gridded EEZ dataset and join with gridded species data

gridded_eez_species <- open_dataset("eez_h3/eez_h3_res7.parquet") %>%
  select(h3 = h3_index, eez = GEONAME) %>%
  as.data.frame() %>%
  left_join(gridded_species, by = "h3") %>%
  filter(!is.na(speciesid))
```

Aggregate by EEZ:

```{r, message=FALSE, warning=FALSE, cache=TRUE}
species_by_eez <- gridded_eez_species %>%
  group_by(speciesid, species, eez) %>%
  summarize() %>%
  arrange(eez, species)

eez_richness_stats <- species_by_eez %>%
  group_by(eez) %>%
  summarize(n_species = n()) %>%
  arrange(eez)

endemics <- species_by_eez %>%
  group_by(speciesid, species) %>%
  summarize(n_eez = n(), eez = first(eez)) %>%
  filter(n_eez == 1) %>%
  select(-n_eez) %>%
  arrange(eez, species)

eez_endemics_stats <- endemics %>%
  group_by(eez) %>%
  summarize(n_endemics = n())
```

## Export lists

```{r, message=FALSE, warning=FALSE, cache=TRUE}
write.csv(endemics, "output/endemics.csv", row.names = FALSE, na = "")
```

## Export statistics

```{r, message=FALSE, warning=FALSE, cache=TRUE}
stats <- eez_richness_stats %>%
  left_join(eez_endemics_stats, by = "eez")

write.csv(stats, "output/stats.csv", row.names = FALSE, na = "")
```

## Visualize

```{r, message=FALSE, warning=FALSE, cache=TRUE}
library(sf)
library(ggplot2)
library(viridis)
library(rnaturalearth)
library(rnaturalearthdata)

sf_use_s2(FALSE)

eezs <- read_sf("eez_gpkg/eez_v11.gpkg") %>%
  select(GEONAME) %>%
  st_simplify(0.05, preserveTopology = FALSE) %>%
  left_join(stats, by = c("GEONAME" = "eez"))

world <- ne_countries(scale = "medium", returnclass = "sf")
```

### Richness

```{r, message=FALSE, warning=FALSE, out.width="100%", fig.width=12, fig.height=8, cache=TRUE}
ggplot() +
  geom_sf(data = eezs, aes_string(fill = "n_species", color = "n_species", geometry = "geom"), lwd = 0.04) +
  scale_color_viridis(option = "inferno", na.value = "white", name = "Number of species", trans = "log10", guide = "none") +
  scale_fill_viridis(option = "inferno", na.value = "white", name = "Number of species", trans = "log10", labels = scales::comma) +
  geom_sf(data = world, fill = "#dddddd", color = "#888888", lwd = 0.1) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(), 
    panel.background = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(2, "cm")
  ) +
  xlab("") + ylab("") +
  coord_sf(crs = "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs" )
```

### Endemics

```{r, message=FALSE, warning=FALSE, out.width="100%", fig.width=12, fig.height=8, cache=TRUE}
ggplot() +
  geom_sf(data = eezs, aes_string(fill = "n_endemics", color = "n_endemics", geometry = "geom"), lwd = 0.04) +
  scale_color_viridis(option = "inferno", na.value = "white", name = "Number of endemics", trans = "log10", guide = "none") +
  scale_fill_viridis(option = "inferno", na.value = "white", name = "Number of endemics", trans = "log10", labels = scales::comma) +
  geom_sf(data = world, fill = "#dddddd", color = "#888888", lwd = 0.1) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(), 
    panel.background = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(2, "cm")
  ) +
  xlab("") + ylab("") +
  coord_sf(crs = "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs" )
```

## Export shapefile

```{r, message=FALSE, warning=FALSE, cache=TRUE}
write_sf(eezs, "output/eezs.gpkg")
```
